---
title: "Transforming Data, Part A"
author: "Aaron Kessler"
date: ""
output: html_document
---

Load the packages we'll need

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)

```

Toy dataset to use, created manually with a tibble  

```{r}
pollution <- tibble(city=c("New York", "New York", "London", "London", "Beijing", "Beijing"),
               size=c("large", "small", "large", "small", "large", "small"),
               amount=c(23, 14, 22, 16, 121, 56))

```

*What are "tibbles"...?*    
 
They're dataframes, with some additional tidyverse-infused features. Returns more readable output in the console.

Let's see the data we just created, you'll see how a tibble view differs in the console

```{r}
pollution

```

Since there are only a handful of rows, a bit harder to see - let's try the built-in "iris" data



```{r}
#(not run)
#iris

```


That was a lot of output! Can limit rows with head()



```{r}
head(iris)

```


But let's see how tibbles differ in their output


```{r}
as_tibble(iris)

```



### FILTERING AND SORTING 

The tidyverse's *dplyr* provides intuitive functions for exploring and analyzing dataframes.    
Let's go back to our little pollution dataset

```{r}
pollution

```

Show me only the ones with a "large" size

```{r}
filter(pollution, size == "large")

```

Show me only the ones where the city is London

```{r}
filter(pollution, city == "London")

```

For numeric values, you can use boolean operators 

```{r}
filter(pollution, amount > 20)
filter(pollution, amount <= 22)

```

Now, let's try filtering based on two different variables.  
*(note the comma separating the filtering terms)*

```{r}
filter(pollution, amount > 20, size == "large")

```

This can still get a little confusing once you wind up with larger amounts of steps to string together.    
    
Enter a glorious feature of the tidyverse: **the PIPE** `%>%`    
    
The "pipe" (shortcut is CTRL/CMD + SHIFT + M) allows you to chain together commands  
 
Watch this, and see how much easier it becomes for a human to think through (and read later!)

```{r}
pollution %>% 
  filter(size == "large")

```

Voila! So what just happened there?  
 
Think of %>% as the equivalent of "and then do this"...  

It takes the result and then applies something new to it, in sequential order  

This becomes easy to see when we add new functions - so let's talk about sorting with `arrange()`

```{r}

pollution %>% 
  arrange(amount)

```

To sort by highest value, add `desc()`

```{r}
pollution %>% 
  arrange(desc(amount))

```

Now let's go back to our filtering and add arranging, too

```{r}
pollution %>% 
  filter(size == "large") %>% 
  arrange(desc(amount))

```

Add another filter criteria

```{r}
pollution %>% 
  filter(size == "large", amount < 100) %>% 
  arrange(desc(amount))

```

This can be formatted this way as well, if it's even easier for you to read.  
Let's add another filter criteria

```{r}
pollution %>% 
  filter(size == "large", 
         amount < 100) %>% 
  arrange(desc(amount))

```

Think about what we just did here.  
 
You can read the code and it intuitively makes sense. 
Each step sequentially listed and executes in order.  

One more thing - what if we don't want all the columns? Just some.  
This happens all the time.

Dplyr makes this easy using `select()`

```{r}

pollution %>% 
  select(city, amount)

```
  
You can pull out just certain variables as well.  
This results in the same thing as above.

```{r}
pollution %>% 
  select(-size)

```


### PRESIDENTIAL CANDIDATE TRIPS  
 
Let's take a look at some more intersting data now and try out some of these methods.  
Load in data of prez candidate campaign trips between midterms and end of Jan. 

```{r}
events <- readRDS("events_saved.rds")

```

Let's take a look at what we've got

```{r}
events

```

Even easier to see a dataset with `View()`  
Click on its name under the environment tab in upper right, or:

```{r}
View(events)

```

Can also pipe the results of a chain if we wanted to:

```{r}
events %>% 
  view()

```

Can you think of when we might find ourselves wanting to do that? (hint: think big)  
  
Now let's try out some of our filtering and arranging techniques.  
  
Show all events in Iowa:

```{r}
events %>% 
  filter(state == "IA")

```

Has Kamala Harris been to Iowa?

```{r}
events %>% 
  filter(state == "IA",
         cand_lastname == "Harris")

```

What about another candidate

```{r}
events %>% 
  filter(state == "IA",
         cand_lastname == "Gillibrand")

```

#### Let's talk about **date-specific** stuff.    
If I have a properly formatted date in a dataframe, can I sort by it? *Yes.*

```{r}
events %>% 
  filter(state == "IA") %>% 
  arrange(desc(date))

```

What if I want to pull out only certain ranges of dates? *Several approaches.*   

Specifiying a specific date using `as.Date()`

```{r}
events %>% 
  filter(date < as.Date("2018-12-31"))

```

Take advantage of the **LUBRIDATE** package - a tidyverse package specifically for dates  
*Note: lubridate needs to be called separately at the top with library(lubridate) - it doesn't yet load with library(tidyverse)*    
  
Now watch what we can do:

```{r}
events %>% 
  filter(year(date) == 2018)

# Just events in January 2019

events %>% 
  filter(year(date) == 2019,
         month(date) == 1)

# Events earlier than Dec 2018

events %>% 
  filter(year(date) == 2018,
         month(date) < 12)

```

Also allows us to do things like, "I only want to see events the *first week of every month*" 

```{r}
events %>% 
  filter(day(date) <= 7)

```

Who's visiting Iowa the first week of a month? 

```{r}
events %>% 
  filter(day(date) <= 7,
         state == "IA")

```

This is helpful but let's say you're doing this all the time.  
It may be easier to create new columns to hold these values.
    
    
Brings us to another key function of dpylr/tidyverse - **MUTATE**

### ADDING COLUMNS WITH MUTATE  
  
To add a column, you give it a name, then a single equal sign (=), then define what's in it.  
Test example:

```{r}
events %>% 
  mutate(mycolumn = "hi there")

```


```{r}
events %>% 
  mutate(electioncycle = 2020)

```

Now let's try adding our date-related columns.  First we'll try year.

```{r}
events %>% 
  mutate(year = year(date))

```

We can add multiple columns as part of one mutate call. Let's do year, month and day in one swoop.

```{r}
events %>% 
  mutate(year = year(date),
         month = month(date),
         day = day(date))

```

This is a good time to remind ourselves that if we want to save our new columns, need to *create new object* or *overwrite*

```{r}
events <- events %>% 
  mutate(year = year(date),
         month = month(date),
         day = day(date))

```

Now we can use our new columns to filter

```{r}
events %>% 
  filter(year == 2019,
         month == 1)

```

Show me just Kamala's events in January


```{r}
events %>% 
  filter(year == 2019,
         month == 1,
         cand_lastname == "Harris")

```


### GROUPING AND AGGREGATING  
 
 
Being able to aggregate our campaign trips would be helpful at this point, right?  
Let's get into how to do it using the tidyverse and dplyr's `group_by()` and `summarise()` 
  
Have you all grouped before in other languages? In base R itself?  Let's discuss.  
  
Grouping to see how many trips each candidate have been on in our data  
Getting used to `n()`

```{r}
events %>% 
  group_by(cand_lastname) %>% 
  summarise(n())

```

now let's add arrange to see who has the most trips

```{r}
# (not run)  
# events %>% 
#   group_by(cand_lastname) %>% 
#   summarise(n()) %>% 
#   arrange(n)

```

hmm - what's going on here? Look closely and see what the generated count column is called

```{r}
# events %>% 
#   group_by(cand_lastname) %>% 
#   summarise(n()) %>% 
#   arrange("n()")

```

That doesn't work either. What about this:

```{r}
events %>% 
  group_by(cand_lastname) %>% 
  summarise(n()) %>% 
  arrange()

```

Ah - so that sort of works? But not really, how do we get desc

```{r}
# events %>% 
#   group_by(cand_lastname) %>% 
#   summarise(n()) %>% 
#   arrange(desc)

```

Oy - this is getting frustrating. How do we solve?  
By doing this: giving the new column a name of our own.  
Check it out:

```{r}
events %>% 
  group_by(cand_lastname) %>% 
  summarise(n = n()) 

```

Now we can do:

```{r}
events %>% 
  group_by(cand_lastname) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n))

```

Bingo.  
We can call the new columnn anything we want. "n" is a common thing for counts,  
but can be anything.

```{r}
events %>% 
  group_by(cand_lastname) %>% 
  summarise(numtrips = n()) %>% 
  arrange(desc(numtrips))

```

Now for the magic  
Because this counting is such a common operation, and because the `n()` becomes a pain to deal with...  
...there is a special shortcut that we can use that collapses everything into one function.  
It's called `count()`.

```{r}
events %>% 
  count(cand_lastname)

# sorting

events %>% 
  count(cand_lastname) %>% 
  arrange(desc(n))

# top states visited

events %>% 
  count(state) %>% 
  arrange(desc(n))

# top months

events %>% 
  count(month) %>% 
  arrange(desc(n))

# top single day for most trips

events %>% 
  count(date) %>% 
  arrange(desc(n))

```

we can also group by **more than one** variable  
which candidates have gone to which states?

```{r}
events %>% 
  count(cand_lastname, state) %>% 
  arrange(state, desc(n))

```

what about the most frequent types of events

```{r}
events %>% 
  count(event_type) %>% 
  arrange(desc(n))

```

Here we're seeing some potentially dirty data that needs cleaning.  
The event types seem to be inconsistently entered.  
How might we standardize them? Let's take a look.  

### USING CONDITIONAL FUNCTIONS
  
A function that returns a vector the same length as the input is called **vectorized**.  
* `ifelse()`  
  
Let's see `ifelse()` in action

```{r}
events %>% 
  mutate(new_type = ifelse(event_type == "event speech", "TEST", event_type)) 

```

Let's talk about what just happened?

Ok, now let's clean a few columns for real:

```{r}
events %>% 
  mutate(new_type = ifelse(event_type == "campaign event", "event", event_type),
         new_type = ifelse(event_type == "campaign events", "event", new_type),
         new_type = ifelse(event_type == "event speech", "speech", new_type)
         ) 

```

This can start to get a little tedious, though.  
Enter `case_when`...

```{r}
events %>%
  mutate(new_type = case_when(
            event_type == "campaign event" ~ "event",
            event_type == "campaign events" ~ "event",
            event_type == "event speech" ~ "speech",
            event_type == TRUE ~ "other"
      ))

```

Of course, you may be asking: wouldn't it be nice if we could standardize...  
...based on certain keywords or patterns?  Instead of spelling out every variation.  
  
The answer is yes. Thanks to "string functions"...!  
  
We'll show a quick example of what that looks like, and then start from the beginning in the next module.  

```{r}
events %>%
  mutate(new_type = case_when(
    str_detect(event_type, "event") ~ "event")
  )

```

We'll next take a closer look at string functions now using the stringr package.  
  
First, are there questions? Let's discuss.






