---
title: "NICAR 2020 - Transforming Data, Mutating and Grouping"
author: ""
date: ""
output: html_document
---

Load the packages we'll need

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)

events <- readRDS("data/events_saved.rds")

```

Another key function of dpylr/tidyverse - **MUTATE**

### ADDING COLUMNS WITH MUTATE  
  
To add a column, you give it a name, then a single equal sign (=), then define what's in it.  
Here's a test example:

```{r}
events %>% 
  mutate(mycolumn = "hi there")

```


```{r}
events %>% 
  mutate(electioncycle = 2020)

```

Now let's try adding our date-related columns.  First we'll try year.

```{r}
events %>% 
  mutate(year = year(date))

```

We can add multiple columns as part of one mutate call. Let's do year, month and day in one swoop.  
*You try it:*

```{r}





```


This is a good time to remind ourselves that if we want to save our new columns, need to *create new object* or *overwrite*

```{r}
events <- events %>% 
  mutate(year = year(date),
         month = month(date),
         day = day(date))

```

Now we can use our new columns to filter

```{r}
events %>% 
  filter(year == 2019,
         month == 1)

```

Show me just Kamala Harris events in January 2019. *You try it:*


```{r}




```


### GROUPING AND AGGREGATING  
 
 
Being able to aggregate our campaign trips would be helpful at this point, right?  
Let's get into how to do it using the tidyverse and dplyr's `group_by()` and `summarise()` 
  
Have you all grouped before in other languages? In base R itself?  Let's discuss.  
  
Grouping to see how many trips each candidate have been on in our data  
Getting used to `n()`

```{r}
events %>% 
  group_by(cand_lastname) %>% 
  summarise(n())

```

now let's add arrange to see who has the most trips

```{r}
# (not run)  
# events %>% 
#   group_by(cand_lastname) %>% 
#   summarise(n()) %>% 
#   arrange(n)

```

hmm - what's going on here? Look closely and see what the generated count column is called

```{r}
# events %>% 
#   group_by(cand_lastname) %>% 
#   summarise(n()) %>% 
#   arrange("n()")

```

That doesn't work either.   
  
Oy - this is getting frustrating. How do we solve?  
By doing this: giving the new column a name of our own.  
Check it out:

```{r}
events %>% 
  group_by(cand_lastname) %>% 
  summarise(n = n()) 

```

Now we can do:

```{r}
events %>% 
  group_by(cand_lastname) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n))

```

Bingo.  
We can call the new columnn anything we want. "n" is a common thing for counts,  
but can be anything. 
 
*You try it this time* - do the same as above but give the counted column your own name:

```{r}






```

Now for the magic  
Because this counting is such a common operation, and because the `n()` becomes a pain to deal with...  
...there is a special shortcut that we can use that collapses everything into one function.  
It's called `count()`.

```{r}
events %>% 
  count(cand_lastname)
```


```{r}
# sorting

events %>% 
  count(cand_lastname) %>% 
  arrange(desc(n))
```

*You try this time* - What are the top states visited in descending order?

```{r}
# top states visited




```

How about the top months for trips?

```{r}
# top months




```

Which single day had the most trips?
```{r}
# top single day for most trips





```

we can also group by **more than one** variable  
which candidates have gone to which states?

```{r}
events %>% 
  count(cand_lastname, state) %>% 
  arrange(state, desc(n))

```

what about the most frequent types of events for each candidate?

```{r}
events %>% 
  count(cand_lastname, event_type) %>% 
  arrange(cand_lastname, desc(n))

```

Here we're seeing some potentially dirty data that needs cleaning.  
The event types seem to be inconsistently entered.  
How might we standardize them? Let's take a look.  

### USING CONDITIONAL FUNCTIONS
  
A function that returns a vector the same length as the input is called **vectorized**.  
* `ifelse()`  
  
Let's see `ifelse()` in action

```{r}
events %>% 
  mutate(new_type = ifelse(event_type == "event speech", "TEST", event_type)) 

```

Let's talk about what just happened?

Ok, now let's clean a few columns for real:

```{r}
events %>% 
  mutate(new_type = ifelse(event_type == "campaign event", "event", event_type),
         new_type = ifelse(event_type == "campaign events", "event", new_type),
         new_type = ifelse(event_type == "event speech", "speech", new_type)
         ) 

```

This can start to get a little tedious, though.  
Enter `case_when`...

```{r}
events %>%
  mutate(new_type = case_when(
            event_type == "campaign event" ~ "event",
            event_type == "campaign events" ~ "event",
            event_type == "event speech" ~ "speech",
            event_type == TRUE ~ "other"
      ))

```

Of course, you may be asking: wouldn't it be nice if we could standardize...  
...based on certain keywords or patterns?  Instead of spelling out every variation.  
  
The answer is yes. Thanks to "string functions"...!  
  
We'll show a quick example of what that looks like, and then start from the beginning in the next module.  

```{r}
events %>%
  mutate(new_type = case_when(
    str_detect(event_type, "event") ~ "event")
  )

```

We'll next take a closer look at string functions later in this sequence in step 04.
  
First, are there questions? Let's discuss.






