---
title: "NICAR 2020 - Transforming Data, Joining"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load the packages we'll need
library(tidyverse) # we'll use the stringr package from tidyverse
library(lubridate)
library(janitor)

  
#load in previous data of prez candidate campaign trips - we'll get back to this in a minute
events <- readRDS("data/events_saved.rds")


```


# Joining Tables 

One of the most powerful things about relational data being able to join tables together.  
  
Let's take a look at how to do that with dpylr and the tidyverse.  
  
First, let's bring in some new data:  

```{r}
key_house_results <- readRDS("data/key_house_results.rds") 
key_house_historical <- readRDS("data/key_house_historical.rds") 

```


What do we have here? Let's take a look and discuss.

```{r}
key_house_results

```


```{r}
key_house_historical

```


This is a common thing to see - tables designed to be joined together based on a common key.  
  
In this case, we have the house district itself as the common key between the two tables.  
  
We'll use dplyr's `inner_join()` function to match the tables based on that column.  
Let's see how that works:  

```{r}

inner_join(key_house_results, key_house_historical)

```

Wait, *that's it*? We haven't even told it *what* to join on.  
  
That's because if the two tables share columns with the same name, it defaults to use them for the join.  
  
If you need to specific which columns in each table to match together, you do it like this:  
  
`inner_join(table1, table2, by = c("table1_columnname" = "table2_columnname"))`  
  
We can also use the pipe to write out a join. It depends on your preference.  

```{r}
key_house_results %>% 
  inner_join(key_house_historical)

```


Now with an explicit mentioning of the column to join:

```{r}
key_house_results %>% 
  inner_join(key_house_historical, by = "house_dist")
```

Remember, if we want to save the results, we need to create a new object

```{r}

joined <- key_house_results %>% 
  inner_join(key_house_historical, by = "house_dist")

```

Let's explore our new joined table using what we've learned so far

```{r}

glimpse(joined)
```


What kinds of questions can we ask, using our dplyr functions? Lots of choices!  
  
Let's start out by getting some aggregate counts.  
How many key races were there?  

```{r}
joined %>% 
  count(keyrace_rating)
```

How many did each party win?

```{r}

joined %>% 
  count(keyrace_rating, winner)

```

How many of those wins were flips?

```{r}
joined %>% 
  filter(!is.na(keyrace_rating)) %>% 
  count(winner, flips)

```

Wait a sec, what was that with the `!is.na()`?  
You can reverse certain functions like `is.na()` - returning only NA rows - by adding a `!` before it.  
Just like with `!=`  
  
Now let's examine just the flipped districts:

```{r}
flipped <- joined %>% 
  filter(flips == "Y") 

flipped
```

*Note: this data is for training purposes only. A few actual results affecting flips aren't reflected here.*  

Now we can start asking some questions about the nature of the flipped districts:

```{r}
flipped %>% 
  count(winner)

```

Quite a lopsided result in favor of the Dems.  
How many flipped districts were above vs. below the national average pct of college grads:

```{r}
flipped %>% 
  count(winner, pct_college_abovebelow_natl)

```

How many flipped districts were above vs. below the the national median income figure

```{r}

flipped %>% 
  count(winner, median_income_abovebelow_natl)

```

Interesting!  
  
Let's do some calculating.  
What was the *average margin of victory* for Dems in flipped districts?  

```{r}

flipped %>% 
  group_by(winner) %>% 
  summarise(mean(margin))

```

Maybe there are some other variables of which we might want to see averages

```{r}

flipped %>% 
  group_by(winner) %>% 
  summarise(mean(pct_college))

```

Could we do both of them at the same time? We can, like this:

```{r}

flipped %>% 
  group_by(winner) %>% 
  summarise(mean(margin), mean(pct_college))

```

Hmm, this isn't bad but what if we had five columns, or ten?  
Is there an easier way?  
  
Yes, let's talk about *scoped functions*.  
  
### Scoped dplyr functions

The idea behind scoped functions: variations on the dplyr commands we've used, but designed to apply to multiple variables.  
  
They generally end with `'_all`, `_at`, and `_if` ... e.g. `summarise_if()`  
  
Let's take a look back at our election data. We could do something like this:  

```{r}

flipped %>% 
  group_by(winner) %>% 
  summarise(mean(margin), 
            mean(pct_college),
            mean(median_income))

```

Or, we could use a scoped function. Here, we'll use `summarise_at()` - designed for when you know specific columns you want.

```{r}
flipped %>% 
  group_by(winner) %>% 
  summarise_at(vars(margin, pct_college, median_income), mean)

```

Sweet, right? That was a lot easier.  
Notice the use of `vars()` above - this is needed when specifying multiple variables.  
The columns/variables you want go in `vars()`, followed by the function to apply to them.  
  
We can even apply *more than one* function at a time:  

```{r}

flipped %>% 
  group_by(winner) %>% 
  summarise_at(vars(margin, pct_college, median_income), c(avg = mean, med = median))

```

We can also group by more than one variable, like below where we look at the entire set of races not just flips.

```{r}
joined %>% 
  group_by(flips, winner) %>% 
  summarise_at(vars(margin, pct_college, median_income), mean)

```

Notice something a little odd with the results? We're getting some NAs.  
Since `mean()` breaks when there are NA values, we need to fix that.  

```{r}

joined %>% 
  group_by(flips, winner) %>% 
  summarise_at(vars(margin, pct_college, median_income), mean, na.rm = TRUE)

```

We can even create our own custom functions (won't get into that in this session, though).  
  
Now what if we wanted to apply our mean to every column in the data?  

```{r}
flipped %>% 
  group_by(winner) %>% 
  summarise_all(mean)

```

We got a lot of warnings there. What happened?  
You can't calculate a mean on a text column, only a numeric one.  
  
Enter `summarise_if()`.

```{r}

flipped %>% 
  group_by(winner) %>% 
  summarise_if(is.numeric, mean)

```

Now we're talking.  
  
Though if we look closely, there are some columns we may decide aren't we want.  
Let's say we don't want to averages of vote percentages for our analysis.  
We could see if there's a pattern to their names? There is, so we can use a `select()`` helper function.  

```{r}
flipped %>% 
  select(-ends_with("vote_pct")) %>% 
  group_by(winner) %>% 
  summarise_if(is.numeric, mean)

```

Perfect.  

We're not going to get into all the select helper functions, but they are very useful,
You can read more about them at:     https://www.rdocumentation.org/packages/dplyr/versions/0.7.2/topics/select_helpers  
    
Congrats on making it this far!  
  
Let's take a short breather, and then we'll come back to talk about the previously mentioned magic called "string functions."