---
title: "NICAR 2020 - Transforming Data, String Functions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load the packages we'll need
library(tidyverse) # we'll use the stringr package from tidyverse
library(lubridate)
library(janitor)

  
#load in previous data of prez candidate campaign trips - we'll get back to this in a minute
events <- readRDS("data/events_saved.rds")


```

#### String Functions - Using the **stringr** package

 Each function starts with `str_`  

 `str_length()`  figure out length of string   
 `str_c()`  combine strings   
 `str_sub()`     substitute string   
 `str_detect()`     detect string in string   
 `str_match()`     does string match   
 `str_count()`     count strings   
 `str_split()`    split strings   
 `str_to_upper()`    convert string to upper case   
 `str_to_lower()`    convert string to lower case   
 `str_to_title()`    convert the first letter of each word to upper case   
 `str_trim()`    eliminate trailing white space   

 


```{r}
### Let's load this data in
messy <- tibble(name=c("Bill Smith", "jane doe", "John Forest-William"),
      email=c("bsmith@themail.com", "jdoe@themail.com", "jfwilliams$geemail.com"),
      income=c("$90,000", "$140,000", "E8500"),
      phone=c("(203) 847-334", "207-999-1122", "2128487345"),
      activites=c("fishing, sailing, planting flowers", "reading, raising flowers, biking", "hiking, fishing"))

messy
```

 What problems do you see?  
 *Tasks*
 
 1. Split name into First name and Last name  
 2. Convert names to title case  
 3. Create a new variable identifying bad email addresses 
 4. Convert income to a new number in US dollars  
 5. Create a new variable containing area code  
 6. Creating a new variable counting how many activities each person is engaged in  
 7. Break activities into a set of useful dummy codes    
     

**String length**  
 *str_length(string)* counts the number of characters in each element of a string or character vector.

```{r}
x <- c("Bill", "Bob", "William")

str_length(x)
```


**Combine strings**  
`str_c(strings, sep="")`    
It's like the equivalent of =concatenate in Excel.    

```{r}
data <- tibble(place = c("HQ", "HQ", "HQ"),
              id = c("A", "B", "C"),
              number = c("001", "002", "003"))

data
```


```{r}
# We can add a string to each value in the *number* column this way:
data %>% 
  mutate(combined=str_c("Num: ", number))
```


```{r}
# You can pass the variable `collapse` to `str_c()` if you're turning an array of strings into one.
data %>% 
    group_by(place) %>% 
    summarize(ids_combined = str_c(number, collapse="-"))
```

**subset strings**  
`str_sub(strings, start, end)` extracts and replaces substrings

```{r}
x <- "Dr. James"

str_sub(x, 1, 3)

```


```{r}
#you can also use this to replace something in a string position
str_sub(x, 1, 3) <- "Mr."

x
```


```{r}
# Negative numbers count from the right.
x <- "baby"

str_sub(x, -3, -1)

str_sub(x, -1, -1) <- "ies"

x
```


**change case**  
 
`str_to_upper(strings)` is upper case  
`str_to_lower(strings)` is lower case  
`str_to_title(strings)` is title case  

```{r}
x <- c("john smith", "Mary Todd", "BILL HOLLIS")

str_to_upper(x)
str_to_lower(x)
str_to_title(x)
```


**trim strings**  
`str_trim(string)` remove white space at the beginning and end of string

```{r}
x <- c(" Assault", "Burglary ", " Kidnapping ")

str_trim(x)
```


**detect matches**   
`str_detect(strings, pattern)` returns T/F  
  
This is actually one of the most useful functions to use when filtering.  
Acts like a wildcard.  


```{r}
x <- c("Bill", "Bob", "David.Williams")
x
```


```{r}
str_detect(x, "il")

```


### Back to our campaign data 

```{r}
events

```

Now let's use string functions to standardize a few event types

```{r}
events %>%
  select(event_type) %>% 
  mutate(new_type = case_when(
    str_detect(event_type, "speech") ~ "speech")
  )
```


Could there be a problem here?  
How would we handle cells with multiple values?

```{r}
events %>%
  select(event_type) %>% 
  mutate(new_type = case_when(
    str_detect(event_type, "speech") ~ "speech",
    str_detect(event_type, "event") ~ "unspecified event",
    str_detect(event_type, "forum") ~ "town hall",
    str_detect(event_type, "town hall") ~ "town hall",
    str_detect(event_type, ",") ~ "multiple", #here we're looking to the comma which is present in multiples
    )
  )
```

Notice that in the example above, the search for comma comes **first*, not last  
  
We can also use our string functions for filtering.      
Let's see what that might look like:

```{r}
events %>% 
  filter(str_detect(event_type, "event"))

```

That's *kinda* helpful, but is there a column this could be even more useful for?  
  
Examine the descriptions

```{r}
events %>% 
  select(cand_restname, description)

```

What we want to find are descriptions mentioning students

```{r}
events %>% 
  select(cand_restname, description) %>% 
  filter(str_detect(description, "student"))

```

How about anything referencing the NAACP?

```{r}
events %>% 
  select(cand_restname, description) %>% 
  filter(str_detect(description, "NAACP"))

```

Remember: R is *case-sensitive*.     
  
Could an acronym like that possibly cause us trouble?  

```{r}
events %>% 
  select(cand_restname, description) %>% 
  filter(str_detect(description, "Naacp"))


```
    
If so, how might we solve the issue of case sensitivity? 

*You try to solve this one:**

```{r}





```


This method is a good strategy to use almost anytime you're searching in this way  
Even when you don't think you'll need it, you never know.    
Let's look at an example...

```{r}
events %>% 
  filter(str_detect(description, "border"))

# No results. Or are there?

events %>% 
  filter(str_detect(str_to_lower(description), "border"))

# You can also do the reverse for the case, with the same goal.

events %>% 
  filter(str_detect(str_to_upper(description), "BORDER"))

```

YOU MADE IT! 